Soundness Proofs for Iterative Deepening

### 2016-09-01 oopsla-16/proof/assumptions.tex
oopsla proofs: brevity and clarity in intro material

This section documents the assumptions we make about the concurrency model, language model, and test environment,
and discusses some limitations that may arise from these assumptions.

@@

{\bf Maximal state space.}
We assume the model checker has all static/hard-coded PPs enabled during this state space exploration,
and that we are not limited by a pressing CPU budget.
Further, we assume that the static PPs include all lock/unlock/trylock operations on mutexes (or whatever other low-level locks are used) and also all higher-level sync primitives which can cause HB (either directly, or because they are built on top of mutexes).

Using only a subset of PPs or aborting early due to time-out could each ruin our ability to reach the goal interleaving or goal state space.
However, Iterative Deepening aims to test the most important interleavings with the time available,
so in the case of not enough time, our point here is that continuing the current state space fits that goal best.
%An alternate approach would be to accept the malloc-recycle PPs, giving them the lowest possible priority, just in case any CPUs would otherwise be idle with not enough jobs to run.
%In that case, our point is to justify deprioritizing those jobs so aggressively.

@@

{\bf Shared memory thread communication.}
We assume that the only way for two threads' transitions to affect each other's behaviour, should they be reordered,
is through either shared memory or a correctly-instrumented sync API.
Both DPOR and data-race detection rely on this assumption,
as any other way for threads to affect each other's behaviour would invisibly reduce independence and break soundness.
For brevity in these proofs, we refer to all thread communication as shared memory,
and assume that other mechanisms, such as system calls that access the filesystem, could be instrumented to fit the same model.

@@

{\bf Locks are correct.}
Because hybrid data-race detection uses lockset analysis to conclude that many access pairs could not possibly race,
we assume that no preemption during a lock-protected critical section could cause a contending thread to make meaningful progress.
This extends to use of disabling/enabling interrupts during kernel-space testing, which we model as a single global lock,
although we do not model RCU \cite{rcu}.
Our data-race analysis also models the behaviour of r/w locks and 1-initialized semaphores (the latter heuristically), although this is tangential to the proof.

@@

@@

{\bf Halting.}
From a certain perspective, proving convergence is the same as proving completeness,
which should be impossible for any runtime analysis of a Turing-complete language \cite{entscheidungsproblem}.
However, being already limited by practical real-world CPU budgets,
we are already accepting that many tests will time out.
We are concerned with proving the limited case that when \quicksand~{\em does} terminate with all state spaces completed, the verification is sound.
Inversely, during our proof, when we show that Iterative Deepening will eventually converge to an arbitrary buggy interleaving, we assume that no intermediate state spaces contain nontermination bugs.
If they do, for the sake of testing/verification, we are satisfied with finding that bug instead.
In this proof, we assume that the MC's heuristic infinite loop checker has no false negatives;
i.e., it will never get stuck forever in an infinite loop without identifying a bug, while occasionally producing false positives.

@@

{\bf Malloc is a magic black box.}
We assume the malloc implementation is correct (e.g., it won't double-allocate blocks), although we don't assume any implementation details such as a tendency to reuse blocks or allocate adjacent ones.
%particular allocation pattern regarding adjacency/coalescing/reuse.
In fact, in our experiments we instruct our MC to ignore all potential PPs which might be inserted on malloc's internal mutex;
in essence treating it as a ``magic primitive'', because we are not interested in verifying its implementation.
Furthermore, we configured DPOR to ignore the internal heap metadata accesses
when tracking shared memory conflicts to achieve greater state space reduction
(i.e., if the only consequence of reordering two transitions is malloc returning different addresses, we consider them independent).
%for the purpose of avoiding malloc-only shared memory conflicts.
This is not without consequences; see section~\ref{sec:owned}.

{\bf Sharing heap addresses.}
Finally, we assume that the only way the program can obtain heap addresses is through the return value of malloc().
Because we are testing C programs, any bizarre violations of this assumption are technically possible,
but should you wish to check for bugs like this,
%we would recommend a data-flow analysis which is much cheaper than model checking anyway.
symbolic execution \cite{klee} would be more appropriate.

For Section \ref{sec:proof}, we further assume a malloced block's address cannot be obtained through arithmetic on the address of a different block; in Section \ref{sec:owned} we show how to account for this case by relaxing the previous ``black box'' assumption.

>>>>

This section documents our assumptions about the concurrency model, language model, and test environment,
and discusses some limitations that may arise therefrom.

@@

{\bf Maximal state space.}
We assume the model checker has all
%static/hard-coded
synchronization API
PPs enabled during this state space exploration,
and that we are not limited by a pressing CPU budget.
%Further, we assume that the static PPs include all lock/unlock/trylock operations on mutexes (or whatever other low-level locks are used) and also all higher-level sync primitives which can cause HB (either directly, or because they are built on top of mutexes).
We assume that all synchronization primitives not listed in our system model are built upon them,
or otherwise annotated for the MC to add additional PPs for them.

Using only a subset of PPs or aborting early due to time-out could each ruin our ability to reach the goal interleaving or goal state space.
However, Iterative Deepening aims to test the most important interleavings with the time available,
so in the case of not enough time,
%our point here is that
continuing the current state space fits that goal best.
%An alternate approach would be to accept the malloc-recycle PPs, giving them the lowest possible priority, just in case any CPUs would otherwise be idle with not enough jobs to run.
%In that case, our point is to justify deprioritizing those jobs so aggressively.

@@

{\bf Shared memory thread communication.}
We assume that the only way for two threads' transitions to affect each other's behaviour, should they be reordered,
is through either shared memory or a correctly-instrumented sync API.
Both DPOR and data-race detection rely on this assumption,
as any other way for threads to affect each other's behaviour could invisibly reduce independence and break soundness.
For brevity in these proofs, we refer to all thread communication as shared memory,
and assume that other mechanisms, such as system calls that access the filesystem, could be instrumented to fit the same model.

@@

{\bf Locks are correct.}
Because hybrid data-race detection uses lockset analysis to exclude many candidate access pairs,
%conclude that many access pairs could not possibly race,
we assume that no preemption during a lock-protected critical section could cause a contending thread to make progress.
This extends to use of disabling/enabling interrupts in kernel code, which we model as a single global lock,
although we do not model RCU \cite{rcu}.

@@

We also assume that any unusual locking discipline, such as recursive mutexes or lock hand-off,
is correctly annotated to the data-race analysis.
Our implementation also models the behaviour of r/w locks and 1-initialized semaphores (the latter heuristically),
although this is tangential to the proof.

@@

{\bf Halting.}
From a certain perspective, convergence is the same as completeness,
which should be impossible for any runtime analysis of a Turing-complete language \cite{entscheidungsproblem}.
However, being already limited by practical real-world CPU budgets,
we are already accepting that many tests will time out.
We are concerned with the limited case of when \quicksand~{\em does} terminate with all state spaces completed.
%, the verification is sound.

Inversely, when we prove that Iterative Deepening will eventually find an arbitrary buggy interleaving,
we assume that no intermediate state spaces contain nontermination bugs.
If they do,
%for the sake of testing/verification,
we are satisfied with finding that bug instead.
In this proof, we assume that the MC's heuristic infinite loop checker has no false negatives;
i.e., it will never get stuck forever in an infinite loop without identifying a bug (necessarily accepting some false positives).

@@

{\bf Malloc is a magic black box.}
We assume the malloc implementation is correct (e.g., it won't double-allocate),
although we don't assume any implementation details such as a tendency to reuse blocks or allocate adjacent ones.
%particular allocation pattern regarding adjacency/coalescing/reuse.
In fact,
%in our experiments we instruct our MC to
our implementation ignores all potential PPs arising from malloc's internal mutex;
in essence treating it as a ``magic primitive'', because we are not interested in verifying its implementation.
Furthermore, we configured DPOR to ignore any shared memory conflicts arising from internal heap metadata
%when tracking shared memory conflicts
to achieve greater state space reduction.
(If the only consequence of reordering two transitions is malloc returning different addresses, we consider them independent).
%for the purpose of avoiding malloc-only shared memory conflicts.
This is not without consequences; see section~\ref{sec:owned}.

{\bf Sharing heap addresses.}
Finally, we assume that the only way the program can obtain heap addresses is through the return value of $\mathsf{malloc}$.
Because we are testing C programs, any bizarre violations of this assumption are technically possible,
but should you wish to check for bugs like this,
%we would recommend a data-flow analysis which is much cheaper than model checking anyway.
symbolic execution \cite{klee} would be more appropriate.

In Section \ref{sec:proof}, we further assume a $\mathsf{malloc}$ed block's address cannot be obtained through arithmetic on the address of a different block;
in Section \ref{sec:owned} we account for this case by relaxing the previous ``black box'' assumption.

### 2016-09-01 oopsla-16/proof/di.tex
oopsla proofs: brevity and clarity in intro material

When generating execution traces, the evaluation semantics is parameterized by a set of active PP predicates.
As long as the set contains the synchronization API PPs,
the execution shall switch threads only when a PP predicate holds.

@@

In the main paper, we used ``PP'' largely to denote what we now call ``PP predicates''.
Hence, the ``same'' PP could occur multiple times during an execution.
%for example, {\tt mutex\_lock()} may be called from {\tt foo()} and later again from {\tt bar()}.
In these proofs, we separate such cases into multiple unique PPs: each PP is simply a label denoting the end of a single transition.

@@

\end{definition}
We require the invariant that each transition's instructions are associated with exactly one thread. (That is, the set of PPs always includes all thread switches.)
The set of synchronization API PPs provides this invariant, and all other PP sets in these proofs will be supersets of those.
%We also assume a trace of all memory accesses is available in the representation of transitions. %% It is, now.

@@

Intuitively, MHB expresses the "cannot be reordered with" relation between transitions
("enables" in DPOR terminology \cite{dpor}).
Two transitions $A$ and $B$ of different threads MHB if some synchronization event in $A$ causes $B$ to become runnable while it was previously blocked.
Such synchronization events include $\mathsf{thread\_fork}$, $\mathsf{make\_runnable}$,
and sometimes but not always $\mathsf{mutex\_unlock}$.

@@

In our main paper, we refer to this relation (in conjunction with a lock-set analysis) as Limited HB.

@@

\begin{definition}[Data race]
A shared memory conflict where furthermore:
\begin{itemize}
	\item The intersection of both threads' locksets is empty (i.e., the threads do not hold the same lock during each access), and
	\item The threads' transitions are not related by MHB.
\end{itemize}
\end{definition}

The same as in the paper, we distinguish between data-race {\em candidates} (or {\em potential} data races) and data-race {\em bugs}.
For brevity, we now use ``data race'' to refer both to true races and to potential data-race candidates identified by MHB.
In this proof we are concerned solely with candidates, and whether they can be observed to race or are false positives.
It is up to the MC to decide whether true data races are benign or buggy.

\begin{definition}[False positive data race]
	An apparent data race that cannot be observed in the opposite order from what was actually executed.
\end{definition}

False positives are caused when some data dependency based on some other shared state, invisible to the data-race analysis,
changes some variable values when the threads are reordered, such that the memory addresses no longer collide.

@@

Recent work \cite{sparc-ssm} has proposed hardware techniques for detecting many classes of stale heap pointer accesses, including the one shown in Figure \ref{fig:recycle-bug}.
This approach could be combined with stateless model checking in future work to identify such bugs immediately,
rather than requiring Iterative Deepening to explore new state spaces corresponding to the data race.
However, if the {\tt malloc} call were in thread 1 instead of thread 2, the bug would still be nondeterministic, requiring stateless model checking to expose.

@@

In this section we provide (hopefully) intuitive summaries of both our proof goals.
%for readers not interested in double-checking the proofs' internal structure.

{\bf Intuition for Iterative Deepening convergence.}
In summary, we are proving that when Iterative Deepening saturates the set of data-race PPs,
that set represents every single instruction where a preemption could possibly affect the program's behaviour,
and that completing the associated state spaces is as strong of a verification as testing all possible thread interleavings by preempting anywhere.
A data-race may be hidden in control-flow paths reachable only after preempting on a different data-race,
but the technique's iterative nature will eventually find it.
On the other hand, relying on the soundness of DPOR, preempting on an instruction which is neither a data-race or sync API boundary cannot affect the program's behaviour,
so any such PPs are irrelevant.

@@

{\bf Intuition for malloc-recycle soundness.}
In summary, we are proving that if a malloc-recycle-pattern data race is a true race, rather than a false positive,
then DPOR is guaranteed to ``reorder away the free and re-malloc''.
In other words, DPOR's exploration will eventually interleave threads in such a way that the malloc-recycle pattern will disappear,
while the access pair remains for the data-race detector to find, as shown in Figure~\ref{fig:recycle-goal}.
Hence, in the same state space where the malloc-recycle data race was found, if it's a true race, the same race will also appear without the recycle pattern.
So if that race hides a failure bug (assertion, segfault, etc.), Iterative Deepening will still be led to the necessary preemption point to find that bug.

>>>>

When generating execution traces, the evaluation semantics is parameterized by a set of active PP predicates.
As long as the set contains the synchronization API PPs,
%the execution shall switch threads only when a PP predicate holds.
the thread-switch invariant discussed above will hold.

@@

In the main paper, ``PP'' referred to what we now call ``PP predicates''.
Hence, the ``same'' PP could occur multiple times during an execution.
%for example, {\tt mutex\_lock()} may be called from {\tt foo()} and later again from {\tt bar()}.
In these proofs, we separate such cases into multiple unique PPs:
each PP is simply a label denoting the boundary between two transitions.

@@

\end{definition}
The thread-switch invariant guarantees that each transition's instructions are associated with exactly one thread.
%The set of synchronization API PPs provides this invariant, and all other PP sets in these proofs will be supersets of those.
%We also assume a trace of all memory accesses is available in the representation of transitions. %% It is, now.

@@

Intuitively, MHB expresses when two transitions cannot be reordered
(the ``enables'' relation in DPOR terminology \cite{dpor}).
Two transitions $A$ and $B$ of different threads MHB if some synchronization event in $A$ causes $B$ to become runnable while it was previously blocked.
Such synchronization events include $\mathsf{thread\_fork}$, $\mathsf{make\_runnable}$,
and sometimes but not always $\mathsf{mutex\_unlock}$.

@@

Our main paper refers to this relation (in conjunction with a lock-set analysis) as Limited HB.

@@

\begin{definition}[Data race]
A shared memory conflict where furthermore:
\begin{itemize}
	\item The intersection of both threads' locksets is empty (i.e., the same lock does not protect both accesses), and
	\item The containing transitions are not related by MHB.
\end{itemize}
\end{definition}

The same as in the paper, we distinguish between data-race {\em candidates} (or {\em potential} data races) and data-race {\em bugs}.
For brevity, we now use ``data race'' to refer both to true races and to potential data-race candidates identified by MHB.
%In this proof we are concerned solely with candidates, and whether they can be observed to race or are false positives.
%It is up to the MC to decide whether true data races are benign or buggy.

\begin{definition}[False positive data race]
	An apparent data race that cannot be executed in the opposite order from what was observed.
\end{definition}

False positives are caused when some data dependency based on some other shared state %, invisible to the data-race analysis,
changes some variable values when the threads are reordered, such that the memory addresses no longer collide.

@@

Recent work \cite{sparc-ssm} proposed hardware techniques for detecting many classes of stale heap pointer accesses, including the one shown in Figure \ref{fig:recycle-bug}.
Future work could combine this approach with MC to identify such bugs immediately,
rather than requiring Iterative Deepening to explore new state spaces corresponding to the data race.
However, if the {\tt malloc} call were in thread 1 instead of thread 2, the bug would still be nondeterministic, requiring MC to expose.

@@

This section provides (hopefully) intuitive summaries of our proof goals.
%for readers not interested in double-checking the proofs' internal structure.

{\bf Intuition for Iterative Deepening convergence.}
%In summary, we are proving
We will prove
that when Iterative Deepening saturates the set of data-race PPs,
that set represents every instruction where a preemption could possibly affect the program's behaviour
Hence, completing the associated state spaces is as strong a verification as testing all possible thread interleavings by preempting anywhere.
A data-race may be hidden in control-flow paths reachable only after preempting on a different data-race,
but the technique's iterative nature will eventually find it.
On the other hand, relying on the soundness of DPOR, preempting on an instruction which is neither a data-race or sync API boundary cannot affect the program's behaviour,
so any such PPs are irrelevant.

@@

{\bf Intuition for malloc-recycle soundness.}
%In summary, we are proving
We prove that if a malloc-recycle-pattern data race is not a false positive, %a true race, rather than a false positive,
then DPOR %is guaranteed to ``reorder away the free and re-malloc''.
%In other words, DPOR's exploration
will eventually interleave threads in such a way that the malloc-recycle pattern will disappear,
while the access pair remains for the data-race detector to find, as shown in Figure~\ref{fig:recycle-goal}.
Hence, in the same state space where the malloc-recycle data race was found, if it's a true race, the same race will also appear without the recycle pattern.
So if that race can lead to a failure, Iterative Deepening will still be led to the necessary preemption point to find it.
